Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-06-04 PaperLatex/paper.tex
delete extraneous

Nervous systems process and transmit information through electrically-excitable membranes. Conductance-based models are the simplest biophysical representation of an electrically-excitable cell (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studied based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small-circuit networks (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). Additionally, these models provide an approachable framework for understanding salient principles of neuroscience. However, challenges remain in simulating biophysically-realistic neuron models. Conductance-based models are typically high-dimensional with many coupled nonlinear differential equations. Conductances are coupled through the membrane potential, and in multicompartment models, all membrane potentials are coupled. Simulators written in languages like "C", "C++", or "hoc" integrate equations quickly, but these simulators often lack the ease-of-use and interoperability of those written in scientific programming languages (e.g. "Python", "Julia", "MATLAB").

Two major approaches have dominated the design of neuron simulators. Some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) are specified in a compiled language with pre-specified network components. These simulators tend to perform very fast computations with little overhead, but suffer from a steep learning curve. Implementations in more approachable languages and use of graphical interfaces mitigate these drawbacks, but obfuscate the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, many simulators have been designed in popular scientific programming languages to emphasize ease-of-use and flexibility. "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}) parse strings of equations that are specified in the scientific programming language. The equations can be translated into a fast implementation language (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits tremendous flexibility for simulating systems of differential equations, but the syntax tends to be verbose and the hierarchical nature of conductance-based models is not generally reflected in the user-generated code. There is no simulator that combines the advantages of both approaches without sacrificing efficiency, ease-of-use, or clarity.

>>>>

Nervous systems process and transmit information through electrically-excitable membranes. Conductance-based models are the simplest biophysical representation of an electrically-excitable cell (\cite{hodgkinQuantitativeDescriptionMembrane1952}). Studies based on the Hodgkin-Huxley formalism now contribute significantly to mainstream research in small-circuit networks (\cite{marderTheoryMotion1995, prinzComputationalApproachesNeuronal2010, prinzInsightsModelsRhythmic2006}). Additionally, these models provide an approachable framework for understanding salient principles of neuroscience. However, challenges remain in simulating biophysically-realistic neuron models. Conductance-based models are typically high-dimensional with many coupled nonlinear differential equations. Conductances are coupled through the membrane potential, and in multicompartment models, all membrane potentials are coupled. Simulators written in languages like "C", "C++", or "hoc" integrate equations quickly, but these simulators often lack the ease-of-use and interoperability of those written in scientific programming languages (e.g. "Python", "Julia", "MATLAB").

Two major approaches have dominated the design of neuron simulators. First, some simulators, such as "NEURON" (\cite{hinesNEURONSimulationEnvironment1997}) are specified in a compiled language with pre-specified network components. These simulators tend to perform very fast computations with little overhead, but suffer from a steep learning curve. Implementations in more approachable languages and use of graphical interfaces mitigate these drawbacks, but obfuscate the underlying algorithms and parameters (\cite{bretteSimulationNetworksSpiking2007, hinesNEURONPython2009}). In contrast, many simulators have been designed in popular scientific programming languages to emphasize ease-of-use and flexibility. "DynaSim" (\cite{sherfeyDynaSimMATLABToolbox2018}), "ANNarchy" (\cite{vitayANNarchyCodeGeneration2015}), "BRIAN" (\cite{stimbergBrianSecondComing2013}) parse strings of equations that are specified in the scientific programming language. The equations can be translated into a fast implementation language (\cite{stimbergEquationorientedSpecificationNeural2014}). This approach permits tremendous flexibility for simulating systems of differential equations, but the syntax tends to be verbose and the hierarchical nature of conductance-based models is not generally reflected in the user-generated code. Neither approach can maintain efficiency, ease-of-use, or clarity without sacrificing one of these aspects.
